---
title: "P Value Introduction"
author: "Darren Gemoets"
date: "February 20, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# P value introduction

* What is a p value?  What does it measure?

* Are p values reliable?  How are they calculated? 

* How else can statistical significance be calculated?

* How does sample size effect p values? What else effects p values?


## Introductory example

One sample t-test.  Very simple (not often used) but allows us to focus on p value concepts.

Suppose we want to test H0: mu=5 vs Ha: mu not equal 5.  Let's do a simulation.

```{r}
n <- 5    # sample size
mu <- 5   # "true" population mean

for (i in 1:10){
samp <- rnorm(n,mean=mu,sd=10)  # generate some data assuming null is true
mean.samp <- mean(samp)  # calculate the mean

test.stat <- (mean.samp-mu)/sd(samp)

if(test.stat > 0){  ## test stat is in right (upper) tail 
pval <- pnorm(test.stat,lower.tail = FALSE)
} else {
pval <- pnorm(test.stat,lower.tail = TRUE)  
}
print(pval)
}
```

What if the data are skewed?  

```{r}
n <- 5000 # sample size
mu <- 5

for (i in 1:10){
samp <- exp(rnorm(n,mean=mu,sd=1))  # generate some data assuming null is true
mean.samp <- mean(samp)  # calculate the mean

test.stat <- (mean.samp-mu)/sd(samp)

if(test.stat > 0){
pval <- pnorm(test.stat,lower.tail = FALSE)
} else {
pval <- pnorm(test.stat,lower.tail = TRUE)  
}
print(pval)
}

samp <- exp(rnorm(n,mean=mu,sd=1)) 
hist(samp)
```

## Randomization example

We can find p values via simulation.

```{r}
## p value based on simulation
n<- 30
N <- 1000
x.bar <- rep(NA,N)
for (i in 1:N){
  x.bar[i] <- mean(sample(c(0,1),n,replace = TRUE))
  }
hist(x.bar)

## data
p.hat <- 0.5

sum(x.bar>p.hat)/N

```


## Sampling distributions

We need to understand how random sampling behaves.  That is, we know that if we take a sample from a population, that sample is only one out of an infinite number we could have chosen, and that each sample will be different.  This variation in sampling needs to be understood.

To this end, lets consider the data from our NHANES dataset as our *population*.  From the population we'll take some random samples.  Load the NHANES data, take a sample of size 50 from the population.  Find the average age of the sample.

```{r}
NHANES <- read.csv("NHANES.csv")

age.sample <- sample(NHANES$age, 50)
mean(age.sample)
```

What value did you get?  Note that we all got different answers (as expected, we all got a different sample). 

Let's see much the sample mean changes with each new sample.  The following code will take a sample of size 50 from the population of ages, find the mean, and store it in a list.  Once done, we'll make a histogram of all of these sample means.

```{r}
n <- 50  ## sample size 
N <- 500 ## number of simulated samples
mean.list <- rep(NA,N)
for (i in 1:N){
 temp<-sample(NHANES$age,n)    # compute sample
 temp.mean <- mean(temp)        # find the mean
 mean.list[i] <- temp.mean       # store it
}
hist(mean.list)
```

We call this distribution a *sampling distribution* and the bell-shape is called a *normal curve*.  Find the mean and standard deviation of the mean.list (all the sample means), and also make a histogram of the NHANES$age variable and find the mean and standard deviation of all the ages in the population.

```{r}
hist(NHANES$age)
mean(NHANES$age)
sd(NHANES$age)
```

What if the population is skewed?  What happens to the sampling distribution?

```{r}
N<- 500
n <- 3
x.bar <- rep(NA,N)
mu=5
for (i in 1:N){
samp <- exp(rnorm(n,mean=mu,sd=1))  # generate some data assuming null is true
x.bar[i] <- mean(samp)  # calculate the mean
}
hist(x.bar)
```


### Confidence intervals

We can do inference with confidence intervals. 

```{r}
samp <- rnorm(30,mean=5,sd=1)  ## data
t.test(samp, mu=4.5)             ## null is mu=1
```


## Statistical power

R shiny app for power calculations: <https://casertamarco.shinyapps.io/power/>

Power curves.
```{r}
library(pwr)
pwr.t.test(n = 50, d = 1, sig.level = 0.05, type = c("one.sample"), alternative = c("two.sided"))

curve(pwr.t.test(x, d = 1, sig.level = 0.002, type = c("one.sample"), alternative = c("two.sided"))$power,from=0, to=100,ylab="Power")

```






## P values via simulation (I'll fix this up for later)

```{r}
## Frequencies A=0.7, B=0.1, C=0.1, D=0.1

N<-10000
y <- rep(NA,N)
for (i in 1:N){
x<-sample(c('A','B','C','D'),4,replace=TRUE,prob = c(0.7,0.1,0.1,0.1))
y[i]<-paste0(x,collapse = "")

}
sort(table(y),decreasing = TRUE)[1:10]/N
sum(y=="ABCD")/N

```


