---
title: "BMS 793 - Introduction to dplyr"
output: html_document
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
## need to load this data frame.
sleep.study <- read.csv("SleepStudy.csv")   ## make sure you have the path correct
```

## Intro to dplyr and the most important concept: piping

Up to this point we've used "temp" variables to store output for intermediate calculations.  The following example illustrates this.

### Example 1
In the sleep.study dataframe the Gender variable is Male=1, Female=0.  

```{r}
sleep.study.M <- subset(sleep.study, Gender=="1")
sleep.study.M.sorted <- sleep.study.M[order(sleep.study.M$GPA, decreasing = TRUE),]
sleep.study.M.top.ten <- sleep.study.M.sorted[1:10,]
sleep.study.M.top.ten
rm(sleep.study.M.top.ten)

```

It would be nice to have a general format to avoid the need for intermediate (temporary) variables.  In the package dplyr is "piping" operator: %>%.  Here is the above example using piping rather than temporary variables.

### Example 2
```{r}
sleep.study %>% 
  filter(Gender=="1") %>% 
  arrange(desc(GPA)) %>%
  slice(1:10) -> sleep.study.M.top.ten
sleep.study.M.top.ten
## could put on one line: sleep.study %>% filter(Gender=="1") %>% arrange(desc(GPA)) %>%  slice(1:10) -> sleep.study.M.top.ten

```

The advantage of using piping in the previous example is it avoids the cluttered code using temporary variables.  More importantly for us, piping is a fundamental UNIX concept as well.  Let's first work with some basic dplyr fuctions, then practice with piping.


## Basic dplyr functions 

Here are some basic functions.  We'll start with filter(), which is the dplyr version of subset(); they work very similarly

### Example 3

```{r}
# these do the same thing
subset(sleep.study,Gender=="1")
filter(sleep.study,Gender=="1")
```

### Example 4

```{r}
filter(sleep.study, Gender=="0" & GPA > 3)
```


The function arrange() will order a data frame by a column.

### Example 5
```{r}
arrange(sleep.study,GPA)
sleep.study[order(sleep.study$GPA),]
```

The arrange() function will order by one group, the order another group within the first group.

### Example 6

```{r}
arrange(sleep.study, ClassesMissed,GPA)

```

The select() function is another way to extract columns from a data frame.

### Example 7

These all do the same thing.

```{r}
select(sleep.study, Gender, GPA)
sleep.study[,c("Gender","GPA")]
subset(sleep.study,select=c(Gender,GPA))
```

Of the basic dplyr functions, group_by() is the most useful. 

### Example 8

Find the average GPA for M and F.  Note that summarise just calls summary functions on grouped data.
```{r}
sleep.Gender <- group_by(sleep.study, Gender)
summarise(sleep.Gender, mean=mean(GPA))
```

Note that sleep.Gender looks the same as the sleep.study dataframe, but is grouped by Gender.

```{r}
class(sleep.Gender)
```

We can group by more than one variable.

### Example 9

```{r}
sleep.Gdr.clss <- group_by(sleep.study, Gender,ClassYear)
summarise(sleep.Gdr.clss, mean=mean(GPA))

```



## Piping 

Let's start simple.  Piping is another way to express function syntax.  

### Example 7 

Think of the %>% as an arrow inputing the value.
```{r}
log(2)
2 %>% log
2 %>% log()
```

### Example 8

```{r}
sleep.study %>% filter(Gender=="1")
filter(sleep.study,Gender=="1")
```

So far this seems silly.  One advantage is we can string %>% together.

### Example 9

(Pull works like select, but the output is a vector and not a dataframe.) 
```{r}
sleep.study %>% filter(Gender=="1") %>% pull(GPA)  %>% mean

```

### Example 10

Revisit Example 2.
```{r}
sleep.study %>%   
  filter(Gender=="1") %>% 
  arrange(desc(GPA)) %>%
  slice(1:10) -> sleep.study.M.top.ten
sleep.study.M.top.ten

```

## Exercises

### 1

Use filter() to find the following subset of the sleep.study data: ClassesMissed=0, LarkOwl=Lark, Drinks < 5  

```{r}

```

### 2 

Use group_by to find the proportion of missed classes for the following subset: LarkOwl=Owl, Drinks > 5, AverageSleep < 7

```{r}

```

### 3 

Use piping (%>%) to find the summary stats (summary()) of the Drinks variable in the sleep.study dataset.

```{r}

```

### 4 

Use piping to round the mean GPA in the sleep study dataset to one decimal place.

```{r}

```

### 5

Look up the mutate() function.  Use piping to create a new variable Drink.Day that divides the Drinks variable by 7 for each observation.  
