---
title: "BMS_793_NMDS.Rmd"
output: html_document
---

```{r}
#library(ggplot2)
#library(stats)
#library(MASS)
#library(scales)
library(vegan)
#library(devtools)
#install_github("vqv/ggbiplot")
#library(dplyr)
#library(ggbiplot)
```


## Multidimensinoal scaling first.  

MDS is also called principle coordinate analysis.  Here is a good reference: <https://rpubs.com/Shaahin/MDS_1> and another <http://personality-project.org/r/mds.html>.  Let's start with an intuitive example where the distances are literal distances.


```{r}
library(psych)
data("cities")

city.location <- cmdscale(cities, k=2)    #ask for a 2 dimensional solution
#round(city.location,0)        #print the locations to the screen
plot(city.location,type="n", xlab="Dimension 1", ylab="Dimension 2",main ="cmdscale(cities)")    #put up a graphics window
text(city.location,labels=names(cities))     #put the cities into the map

```

```{r}
city.location <- -city.location
plot(city.location,type="n", xlab="Dimension 1", ylab="Dimension 2",main ="cmdscale(cities)")    #put up a graphics window
text(city.location,labels=names(cities))     #put the cities into the map
```


In an nutshell, non-metric multidimensial scaling is multidimensianal scaling where we treat the data as ordinal (rankings) instead of quantitiative.  Let's try NMDS on the cities data.  

```{r}
tst <- monoMDS(cities)  ## from the vegan package
plot(tst)
```

Looks similar, eh?

Let's work though the dune data from the vegan package.  I'm following the example from: <https://jonlefcheck.net/2012/10/24/nmds-tutorial-in-r/> 

```{r}
data("dune")
ord <- metaMDS(dune)
#ordiplot(ord)
ordiplot(ord,type="n")
orditorp(ord,display="species",col="red",air=0.01)
orditorp(ord,display="sites",cex=1.25,air=0.01)
abline(h=0);abline(v=0)
```

Is it a good fit?

```{r}
stressplot(ord)
ord$stress
```

Add some environmental variables.

```{r}

data(dune.env)
attach(dune.env)  #careful!
plot(ord, disp="sites", type="n")
#ordihull(ord, Management, col=1:4, lwd=3)
ordiellipse(ord, Management, col=1:4, kind = "ehull", lwd=3)
#ordiellipse(ord, Management, col=1:4, draw="polygon")
ordispider(ord, Management, col=1:4, label = TRUE)
points(ord, disp="sites", pch=21, col="red", bg="yellow", cex=1.3)

```

Add some more environmental variables.
```{r}
ord.fit <- envfit(ord ~ A1 + Management, data=dune.env, perm=999)
ord.fit
ordiplot(ord, type ="n")
orditorp(ord,display="sites",cex=1.25,air=0.01)
plot(ord.fit)
ordiellipse(ord, Management, col=1:4, kind = "se", lwd=3, conf=0.95)

ordisurf(ord, A1, add=TRUE)

```

If we have time we'll play with the data here: <http://geoffreyzahn.com/nmds-example/>



