---
title: "BMS_793_PCA.Rmd"
output: html_document
---

```{r setup, include=FALSE}
##knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(stats)
library(MASS)
library(scales)
```

## Intro PCA Example

Simple two dimensional example with fake data.  Create and plot some fake bivariate data.


```{r simple}
set.seed(666)  ## a bad seed
dat<-mvrnorm(n = 500, mu=c(0,0), Sigma=matrix(c(2,1.5,1.5,2),nrow=2, ncol=2)) 
plot(dat)
```

Using the `prcomp()` and `princomp()` functions in R to perform principle components analysis.  Here is a link outlining the differences: <http://www.sthda.com/english/articles/31-principal-component-methods-in-r-practical-guide/118-principal-component-analysis-in-r-prcomp-vs-princomp/>

```{r}
#princomp(dat,cor=FALSE)
temp<-prcomp(dat)
#temp.eigen<-eigen(cov(dat))
temp  ## print the principle components.
pca<-princomp(dat, cor=FALSE)
summary(pca)   ## different output
pca$sdev^2/sum(pca$sdev^2)  ## find proportion of variance explained

```

Usually we use PCA to create plots.  The code below adds vectors that show the directions of the two princ. components.  What do you see?

```{r}
par(pty="s")
plot(dat,ylim=c(-4,4),xlim=c(-4,4))
arrows(0, 0, -0.7185, -0.6955,length=0.15,col="red")
arrows(0, 0, -0.6955, 0.7185,length=0.15,col="blue")
       #col = par("fg"), lty = par("lty"), xpd = FALSE)


```

Make a PCA plot.  Both base R and ggplot are below. 

```{r}
pca <- prcomp(dat,scale.=TRUE,center=TRUE)
pca.data = data.frame(pca = pca$x)
prop.pca <- pca$sdev^2/sum(pca$sdev^2)  ## find proportion of variance explained
plot(pca.PC2~pca.PC1,data=pca.data)  ## using base R

ggplot(pca.data) + geom_point(aes(pca.PC1, pca.PC2), size = 2.5) +
  labs(x = paste("PC1 (", percent(prop.pca[1]), ")", sep=""),
       y = paste("PC2 (", percent(prop.pca[2]), ")", sep=""))


```


You try one.  The example below contains uncorrelated data.  Perform PCA.   

```{r}
set.seed(999)
dat<-mvrnorm(n = 500, mu=c(0,0), Sigma=matrix(c(2,0,0,2),nrow=2, ncol=2)) 
par(pty="s")
plot(dat,ylim=c(-4,4),xlim=c(-4,4))

pca <- prcomp(dat)
pca.data = data.frame(pca = pca$x)
prop.pca <- pca$sdev^2/sum(pca$sdev^2)  ## find proportion of variance explained

ggplot(pca.data) + geom_point(aes(pca.PC1, pca.PC2), size = 2.5) +
  labs(x = paste("PC1 (", percent(prop.pca[1]), ")", sep=""),
       y = paste("PC2 (", percent(prop.pca[2]), ")", sep=""))
```

This example already has most of the variability in the x direction.  Can you guess what the princ. comp. decomposition will look like?

```{r}
set.seed(123)
dat<-mvrnorm(n = 500, mu=c(0,0), Sigma=matrix(c(2,0,0,0.1),nrow=2, ncol=2)) 
par(pty="s")
plot(dat,ylim=c(-4,4),xlim=c(-4,4))

pca <- prcomp(dat)#,scale.=TRUE,center=TRUE)
pca.data = data.frame(pca = pca$x)
prop.pca <- pca$sdev^2/sum(pca$sdev^2)  ## find proportion of variance explained

ggplot(pca.data) + geom_point(aes(pca.PC1, pca.PC2), size = 2.5) +
  labs(x = paste("PC1 (", percent(prop.pca[1]), ")", sep=""),
       y = paste("PC2 (", percent(prop.pca[2]), ")", sep=""))
```

### PCA with categorical variable

In practice we use PCA to visualize "clusters".  Here is a very simple example.

```{r}
## make fake data, create data frame with categorical variable
set.seed(123)
dat<-mvrnorm(n = 500, mu=c(0,0), Sigma=matrix(c(2,1.5,1.5,2),nrow=2, ncol=2)) 
datf<-data.frame(x=dat[,1],y=dat[,2],catgry=NA)
## color the categories by location
datf[which(datf$x>0),]$catgry<-"A"
datf[which(datf$x<0),]$catgry<-"B"
datf$catgry <- as.factor(datf$catgry)
## see what it looks like
plot(y~x, data=datf, col=as.integer(datf$catgry))
```

Now do PCA. 
```{r}
pca <- prcomp(datf[,-3],  ## omit categorical variables
              center = TRUE,  ## can use if you want.
              scale. = TRUE) 

prop.pca <- pca$sdev^2/sum(pca$sdev^2)    ## percent variation explained


plot(pca, type="l") # a "scree" plot (we'll discuss later)

dataset = data.frame(grps = datf[,3], pca = pca$x)

ggplot(dataset) + geom_point(aes(pca.PC1, pca.PC2, colour = grps, shape = grps), size = 2.5) +
  labs(x = paste("PC1 (", percent(prop.pca[1]), ")", sep=""),
       y = paste("PC2 (", percent(prop.pca[2]), ")", sep=""))

#grid.arrange(p1, p2)
```


Here is an example where there is no pattern to the colors.  

```{r}
set.seed(123)
dat<-mvrnorm(n = 100, mu=c(0,0), Sigma=matrix(c(2,1.5,1.5,2),nrow=2, ncol=2)) 

datf<-data.frame(x=dat[,1],y=dat[,2],catgry=NA)
datf$catgry<-sample(c("A","B"),size=length(datf$x),replace=TRUE)
datf$catgry <- as.factor(datf$catgry)
plot(y~x, data=datf, col=as.integer(datf$catgry))
```

### Three dimensional example

Try to use what you learned above for a three dimensional example.  Now you will only plot the two principle components that contribute the most variability.


```{r}
dat<-mvrnorm(n = 500, mu=c(0,0,0), Sigma=matrix(c(0.5,0.1,0.1,0.1,3,0.1,0.1,0.1,5),nrow=3, ncol=3))
par(pty="s")
plot(dat[,c(1,2)],ylim=c(-4,4),xlim=c(-4,4))
plot(dat[,c(1,3)],ylim=c(-4,4),xlim=c(-4,4))
plot(dat[,c(2,3)],ylim=c(-4,4),xlim=c(-4,4))
```

```{r}
pca <- prcomp(dat,  ## omit categorical variables
              center = TRUE,  ## can use if you want.
              scale. = TRUE) 

prop.pca <- pca$sdev^2/sum(pca$sdev^2)    ## percent variation explained


plot(pca, type="l") # a "scree" plot (we'll discuss later)

dataset = data.frame(pca = pca$x)

ggplot(dataset) + geom_point(aes(pca.PC1, pca.PC2), size = 2.5) +
  labs(x = paste("PC1 (", percent(prop.pca[1]), ")", sep=""),
       y = paste("PC2 (", percent(prop.pca[2]), ")", sep=""))


```

## Example using the iris petals data

The `iris` data has one categorical variable (Species) and four quantitative variables.  
```{r}
data("iris")
```


Plot of two quant. variables.

```{r}
ggplot(iris) + geom_point(aes(Sepal.Length, Sepal.Width, colour = Species), size = 2.5) 
```

PCA with only the first two quant. variables.

```{r}
pca <- prcomp(iris[,c(1,2)],  ## omit categorical variables
              center = TRUE,  ## can use if you want.
              scale. = TRUE) 

prop.pca <- pca$sdev^2/sum(pca$sdev^2)    ## percent variation explained


plot(pca, type="l") # a "scree" plot (we'll discuss later)

dataset = data.frame(grps = iris[,5], pca = pca$x)

ggplot(dataset) + geom_point(aes(pca.PC1, pca.PC2, colour = grps), size = 2.5) +
  labs(x = paste("PC1 (", percent(prop.pca[1]), ")", sep=""),
       y = paste("PC2 (", percent(prop.pca[2]), ")", sep=""))

```


For you.  Do PCA using all four quantitative variables (just use the code above).

```{r}
pca <- prcomp(iris[,-5],  ## omit categorical variables
              center = TRUE,  ## can use if you want.
              scale. = TRUE) 

prop.pca <- pca$sdev^2/sum(pca$sdev^2)    ## percent variation explained


plot(pca, type="l") # a "scree" plot (we'll discuss later)

dataset = data.frame(grps = iris[,5], pca = pca$x)

ggplot(dataset) + geom_point(aes(pca.PC1, pca.PC2, colour = grps), size = 2.5) +
  labs(x = paste("PC1 (", percent(prop.pca[1]), ")", sep=""),
       y = paste("PC2 (", percent(prop.pca[2]), ")", sep=""))

```


### Another example if time

```{r}
## from https://www.r-bloggers.com/how-to-perform-pca-on-r/
pca<-princomp(USArrests, cor=TRUE)

```

### Another example (if time)

<http://www.sthda.com/english/articles/31-principal-component-methods-in-r-practical-guide/118-principal-component-analysis-in-r-prcomp-vs-princomp/>

