---
title: "BMS_793_PCA_part 2.Rmd"
output: html_document
---

```{r}
library(ggplot2)
library(stats)
library(MASS)
library(scales)
library(devtools)
install_github("vqv/ggbiplot")
library(dplyr)
library(ggbiplot)
```


## PCA with iris data using datacamp tutorial

Borrowing from <https://www.datacamp.com/community/tutorials/pca-analysis-r>

Basic plot.  I also commuted the correlation in the data, and after the PCA has been performed.
```{r}
iris.pca <- prcomp(iris[,-5], center = TRUE,scale. = TRUE)

summary(iris.pca)
str(iris.pca)
cor(iris[,-5])
cor(iris.pca$x)

ggbiplot(iris.pca)
ggbiplot(iris.pca, labels=rownames(iris))  ## not meaningful

```


Add colors for Species. 
```{r}
#mtcars.country <- c(rep("Japan", 3), rep("US",4), rep("Europe", 7),rep("US",3), "Europe", rep("Japan", 3), rep("US",4), rep("Europe", 3), "US", rep("Europe", 3))

ggbiplot(iris.pca,ellipse=TRUE,  groups=iris$Species)  ##labels=rownames(mtcars), 
ggbiplot(iris.pca,ellipse=TRUE,choices=c(3,4),groups=iris$Species) ##labels=rownames(mtcars)
ggbiplot(iris.pca,ellipse=TRUE,circle=TRUE, groups=iris$Species)  ##labels=rownames(mtcars), 


```


```{r}
ggbiplot(iris.pca,ellipse=TRUE,obs.scale = 1, var.scale = 1, groups=iris$Species)  ## labels=rownames(mtcars),
ggbiplot(iris.pca,ellipse=TRUE,obs.scale = 1, var.scale = 1, var.axes=FALSE,    groups=iris$Species) ##labels=rownames(mtcars),

ggbiplot(iris.pca,ellipse=TRUE,obs.scale = 1, var.scale = 1,  groups=iris$Species) + ##labels=rownames(mtcars),
  scale_colour_manual(name="Origin", values= c("forest green", "red3", "dark blue"))+
  ggtitle("PCA of Iris dataset")+
  theme_minimal()+
  theme(legend.position = "bottom")
```


Add new data point.  Both by including in the data first then doing PCA, and also by adding to the PCA after the fact.

```{r}
new.flower <- c(15,10,7,12, NA)

irisplus <- rbind(iris, new.flower)
irisplus$Species[151] <- "virginica"
#mtcars.countryplus <- c(mtcars.country, "Jupiter")

irisplus.pca <- prcomp(irisplus[,-5], center = TRUE,scale. = TRUE)

ggbiplot(irisplus.pca, obs.scale = 1, var.scale = 1, ellipse = TRUE, circle = FALSE, var.axes=TRUE, groups=irisplus$Species)+ ##labels=rownames(mtcars),
  scale_colour_manual(name="Origin", values= c("forest green", "red3", "violet", "dark blue","black"))+
  ggtitle("PCA of mtcars dataset, with extra sample added")+
  theme_minimal()+
  theme(legend.position = "bottom")
```



```{r}
s.sc <- scale(t(new.flower[-5]), center= iris.pca$center)
s.pred <- s.sc %*% iris.pca$rotation    ## or t(iris.pca$rotation) %*% t(s.sc)


iris.plusproj.pca <- iris.pca
iris.plusproj.pca$x <- rbind(iris.plusproj.pca$x, s.pred)


ggbiplot(iris.plusproj.pca, obs.scale = 1, var.scale = 1, ellipse = TRUE, circle = FALSE, var.axes=TRUE, groups=irisplus$Species)+
  scale_colour_manual(name="Origin", values= c("forest green", "red3", "violet", "dark blue","black"))+
  ggtitle("PCA of mtcars dataset, with extra sample projected")+geom_point(aes(x=s.pred[1],y=s.pred[2]),color="orange")
  theme_minimal()+
  theme(legend.position = "bottom")

```



## PCA with yeast data

Just an attempt. These data seem very noisy.

```{r}
yeast.data <- read.csv("./yeast_data/SC_expression.csv",row.names = 1)
yeast.covar <- read.csv("./yeast_data/conditions_annotation.csv")
yeast.pca <- prcomp(yeast.data[,20:30], center = TRUE,scale. = TRUE)
ggbiplot(yeast.pca,var.axes=FALSE)
ggbiplot(yeast.pca, var.axes=FALSE,labels=rownames(yeast.data))

```


Seem to be outliers.  Try to find and remove outliers. 
```{r}
grep("^__align",rownames(yeast.data))
grep("^__no",rownames(yeast.data))
grep("^__ambig",rownames(yeast.data))
yeast.new <- yeast.data[-c(696,4381,5114,5664),]
yeast.pca <- prcomp(yeast.new, center = TRUE,scale. = TRUE)
ggbiplot(yeast.pca,var.axes=FALSE,labels=rownames(yeast.new))
#ggbiplot(yeast.pca, var.axes=FALSE,labels=rownames(yeast.data))

```


Maybe try working with proportions?
```{r}
## calc the proportions for each observation
#yeast1 <- aggregate(.~X,data=yeast.data, FUN = sum,na.rm=TRUE, na.action=NULL)
temp <- colSums(yeast.data[,-1])
yeast.data.prop <- yeast.data
yeast.data.prop[,-1] <- sweep(yeast.data.prop[,-1],MARGIN=2,temp,"/")
yeast.new <- yeast.data.prop[apply(yeast.data.prop[, -1], MARGIN = 1, function(x) any(x > 0.1)), ]

#x<-which(yeast.data.prop[,-1]<0.5)
#length(x)
yeast.pca <- prcomp(yeast.new[,-1], center = TRUE,scale. = TRUE)
ggbiplot(yeast.pca,var.axes=FALSE)
```


Try working with a subset of the columns.
```{r}
yeast.sub <- yeast.data.prop[,yeast.covar$ID[c(76,80,88)]]
yeast.pca <- prcomp(yeast.sub[,-1], center = TRUE,scale. = TRUE)
ggbiplot(yeast.pca,scale.obs=2,var.axes=TRUE)
cor(yeast.sub)
```






